---
title: Association Rules Analysis
author: Noah MacGillivray
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
library(readxl)
library(ggplot2)
library(knitr)
library(arules)
library(arulesViz)
library(ggmosaic)
library(mosaic)
library(vcd)
library(htmlwidgets)
```

**BIT-446 Data Mining Foundations**

# Introduction
This is a Data Mining Foundations project exploring the economic outlook for the state
of Montana using Association Rules and the Apriori Algorithm.

## Read in data from excel spreadsheet. 
This chunk reads in the first 210 rows (eliminating the blank line and note at the 
end), TODO 

```{r}
# Read only the first 210 rows
polling <- read_excel("BIT-446-RS-T3-Poll2.xlsx", n_max = 210)

# Replace values ending with asterisk with NA
for (col in names(polling)) {
  # Find values that end with asterisk (like "POL*", "STAT*")
  asterisk_rows <- grepl("\\*$", polling[[col]])
  asterisk_rows[is.na(asterisk_rows)] <- FALSE  # Handle NAs separately
  polling[[col]][asterisk_rows] <- NA
}

# 3. Rename columns for better visualization
names(polling) <- c("Age_Group", "Gender", "Income_Level",
                    "Political_Affiliation", "Geographic_Area",
                    "Financial_Status", "State_Economic_Outlook")
write.csv(polling, "cleaned_polling_data.csv", row.names = FALSE)

summary(polling)
head(polling, 10)
tail(polling, 10)
```

## Mosaic Plots 

```{r}
# Disable warnings
# options(warn = -1)

# Get all variable names except STAT
other_vars <- names(polling)[names(polling) != "State_Economic_Outlook"]

# Define variable-specific reference information
var_info <- list(
  "Age_Group" = "AGE: 1 under 35, 2 35-54, 3 55 and over",
  "Gender" = "GENDER: 0 male, 1 female",
  "Income_Level" = "INC: 1 under $20K, 2 20-35$K, 3 over $35K",
  "Political_Affiliation" = "POL: 1 Democrat, 2 Independent, 3 Republican",
  "Geographic_Area" = "AREA: 1 Western, 2 Northeastern, 3 Southeastern Montana",
  "Financial_Status" = "FIN: 1 worse, 2 same, 3 better than a year ago"
)

# Create separate 2D mosaic plots with STAT on y-axis
for (var in other_vars) {
  # Create mosaic plot with STAT on y-axis (put STAT last in formula)
  mosaic(as.formula(paste("~", var, "+ State_Economic_Outlook")),
         data = polling,
         main = paste(var, "vs State_Economic_Outlook"),
         xlab = var,
         ylab = "State Economic Outlook",
         labeling = labeling_values,
         gp = gpar(fill = c("lightblue", "lightcoral", "lightgreen"))) # Add colors
  
  # Print variable-specific reference information
  if (var %in% names(var_info)) {
    cat(var_info[[var]], "\n")
  }

  # Print STAT reference
  cat("STAT: 0 better, 1 not better than a year ago\n\n")
}

# Re-enable warnings
# options(warn = 0)
```

## Prepare dataframe for item set and association rules analysis. 
Note: we have already accounted for the frames without information which include and '*'. 
The following continued the data cleaning by converting each column from character/text to 
a factor type. It then removes the rows which have empty frames (NA).

```{r}
# Quick factor conversion for association rules (the key hint!)
data_factors <- data.frame(lapply(polling, as.factor))

# Clean any NA or missing values if needed.
data_factors <- data_factors[complete.cases(data_factors), ]

# Show results of factor conversion and cleaning NA values. 
head(data_factors, 10)

```

## Create itemsets using arules package.
Support = 0.10 | Confidence = 0.70
minlen = 2 | maxlen = 10

```{r}
# Convert to transactions object for arules analysis
transactions <- as(data_factors, "transactions")

# Generate frequent itemsets
cat("=== GENERATING FREQUENT ITEMSETS ===\n")
frequent_itemsets <- apriori(transactions,
                           parameter = list(support = 0.10,
                                          confidence = 0.70,
                                          minlen = 2,
                                          maxlen = 10,
                                          target = "frequent itemsets"))

cat("Frequent Itemsets Summary:\n")
summary(frequent_itemsets)
cat("\n")

# Inspect frequent itemsets (sorted by support)
cat("=== FREQUENT ITEMSETS (sorted by support) ===\n")
cat("\n")

# Sort the itemsets by support for easy viewing. 
frequent_itemsets_sorted <- sort(frequent_itemsets, by = "support", decreasing = TRUE)

# Package conflict; had to reload the arules package. 
arules::inspect(frequent_itemsets_sorted)

```

## Create interactive inspection 

```{r}
# Create interactive inspection
cat("=== CREATING INTERACTIVE HTML EXPORT ===\n")
itemsets_interactive <- inspectDT(frequent_itemsets_sorted)

# Display in RStudio viewer/browser
itemsets_interactive

# Export to HTML file
htmlwidgets::saveWidget(itemsets_interactive, 
                       file = "item_sets.html",
                       selfcontained = TRUE)
```

## Association rules analysis using arules package

```{r}
# Create association rules with STAT as consequent
rules <- apriori(transactions,
                parameter = list(
                  supp = 0.05,      # Experiment with these values
                  conf = 0.60,      # Experiment with these values
                  minlen = 1,
                  maxlen = 10,
                  target = "rules"
                ),
                appearance = list(
                  rhs = c("State_Economic_Outlook=STAT0", "State_Economic_Outlook=STAT1"),
                  default = "lhs"
                ))

# Filter rules with lift > 1.3
high_lift_rules <- subset(rules, lift > 1.3)

arules::inspect(high_lift_rules)
```